A Linear Solution of the Linearity Challenge

### 2025-12-26 paper/Examples.lagda.tex
revision

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Function using (_$_)
open import Data.Sum using (inj₁; inj₂)
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Data.Fin using (zero; suc; #_)
open import Data.List.Base using ([]; _∷_; [_])
open import Relation.Unary

@@

In this section we revisit the processes discussed in \Cref{ex:booleans,ex:echo}
and show their encoding in our formalisation. The encoding of $\Bool$ is
straightforward.

\begin{code}
𝔹 : Type
𝔹 = 𝟙 ⊕ 𝟙
\end{code}

The boolean constants are encoded thus:
\begin{code}
True : Proc [ 𝔹 ]
True = select (ch ⟨ < ≫ ⟩ inj₁ (close ch))

@@

We can take advantage of the host language for programming higher-order
processes. For example, we can define a conditional process thus:

@@

The function \AgdaFunction{\$} (defined in Agda's standard library) is just a
visible function application operator. We use it here as ``separator'' to save a
few parentheses.
%
Note the use of \AgdaInductiveConstructor{cut}s for combining a process that
produces a boolean value with a process that consumes a boolean value and the
use of \AgdaInductiveConstructor{link} for copying a boolean value from one
channel to another. 
%
Here we have chosen to define the disjunction \AgdaFunction{||} from the
conjunction \AgdaFunction{\&\&} and negation \AgdaFunction{!!} using De Morgan's
laws.

@@

For the encoding of the echo server (\Cref{ex:echo}), recall that the server is
polymorphic in the sense that the type $X$ of the value it receives is
arbitrary. The notation $\AgdaFunction{\#}~n$ is defined in Agda's standard
library and creates an element of \AgdaDatatype{Fin} from the natural number
$n$.

\begin{code}
Echo : let X = var (# 0) in
       Proc [ `! (`∀ (X ⅋ (dual X ⊗ 𝟙))) ]
Echo = curry∗ server ch (< ≫)
             ( un-[]
             , curry∗ all ch (< ≫) λ X →
               curry∗ join ch (< ≫) $
               curry∗ (curry∗ fork ch (< ≫)) (curry∗ link ch (< > •) ch) (< ≫) $
               close ch )
\end{code}

We can test the server by connecting it with a \AgdaInductiveConstructor{client}
that sends true to it.

\begin{code}
Echo-True : Proc [ 𝔹 ]
Echo-True = curry∗ cut Echo ≫ $
            curry∗ client ch (< ≫) $
            curry∗ (ex {_} {dual 𝔹}) ch (< ≫) $
            curry∗ (curry∗ fork ch (< ≫)) True ≫ $
            curry∗ join ch (< ≫) $
            curry∗ wait ch (< ≫) $
            curry∗ link ch (< > •) ch
\end{code}

If we ask Agda to evaluate this process we obtain \AgdaFunction{True}, as
expected.

From these examples it should be clear that writing \Calculus processes in Agda
is not entirely straightforward and requires a little effort. The lack of names
(for channels) clearly obfuscates the code, but the context splittings that must
be provided are usually small and fairly easy to understand. Most importantly,
the use of \AgdaFunction{≫} helps truncating context splittings as soon as the
seeked channel has been reached.
%
Either way, the formalisation is meant to study the meta-theory of \Calculus
rather than implementing a user-friendly DSL for it.

>>>>

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Function using (_$_)
open import Data.Sum using (inj₁; inj₂)
open import Data.Product using (_×_; _,_; ∃; ∃-syntax; curry)
open import Data.Fin using (zero; suc; #_)
open import Data.List.Base using ([]; _∷_; [_])
open import Relation.Unary

@@

In this section we revisit and expand the processes discussed in
\Cref{ex:booleans,ex:echo} and show their encoding in our formalisation. The
encoding of $\Bool$ is straightforward

\begin{code}
𝔹 : Type
𝔹 = 𝟙 ⊕ 𝟙
\end{code}
and the boolean constants are encoded thus:
\begin{code}
True : Proc [ 𝔹 ]
True = select (ch ⟨ < ≫ ⟩ inj₁ (close ch))

@@

We take advantage of the host language for programming higher-order processes.
For example, we can define a conditional process thus:

@@

The function \AgdaFunction{\$} (defined in Agda's standard library) is just a
visible function application operator. We use it as a ``separator'' so that we
can flatten deeply nested expressions and save a bunch of parentheses.
%
We have chosen to define the disjunction \AgdaFunction{||} from the conjunction
\AgdaFunction{\&\&} and negation \AgdaFunction{!!} using De Morgan's laws.

@@

For the encoding of the polymorphic echo server (\Cref{ex:echo}), we start by
encoding its type $\OfCourse(\forall\X.\dual\X \Par (X \Ten \One))$:

\begin{code}
EchoServerT : Type
EchoServerT = `! (`∀ (rav (# 0) ⅋ (var (# 0) ⊗ 𝟙)))
\end{code}

The notation $\AgdaFunction{\#}~n$ is defined in Agda's standard library and
creates an element of \AgdaDatatype{Fin} from the natural number $n$. In
\AgdaFunction{EchoServer}, we use it to represent the de Bruijn index of the
type variable $X$.
%
We now encode the server
\begin{code}
Server : Proc [ EchoServerT ]
Server = curry (curry∗ server ch (< ≫)) un-[] $
         curry∗ all ch (< ≫) λ X →
         curry∗ join ch (< ≫) $
         curry∗ (curry∗ fork ch (< ≫)) (curry∗ link ch (< > •) ch) (< ≫) $
         close ch
\end{code}
and the client that sends true to it
\begin{code}
Client : Proc (dual EchoServerT ∷ 𝔹 ∷ [])
Client = curry∗ client ch (< ≫) $
         curry∗ (ex {_} {𝔹}) ch (< ≫) $
         curry∗ (curry∗ fork ch (< ≫)) True ≫ $
         curry∗ join ch (< ≫) $
         curry∗ wait ch (< ≫) $
         curry∗ link ch (< > •) ch
\end{code}

To test these definitions, we define a main process that connects server and
client
\begin{code}
Main : Proc [ 𝔹 ]
Main = curry∗ cut Client (< •) Server
\end{code}
and then ask Agda to normalize \AgdaFunction{Main}, which yields
\AgdaFunction{True} as expected.
